---
title: "Predicting the Manner of Which Weight Lifting Exercises are Performed"
author: "coursera-2015"
date: "Saturday, June 20, 2015"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

A test was conducted where 6 participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways:

 - Class A: exactly according to the specification
 - Class B: throwing the elbows to the front
 - Class C: lifting the dumbbell only halfway
 - Class D: lowering the dumbbell only halfway
 - Class E: throwing the hips to the front
 
More information on this test is available from the [website](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

## Goal

In this project, my goal is to use data from accelerometers on the belt, forearm, arm, and dumbbell of the 6 participants to predict the manner in which they did the exercise. 

## Load caret Library

```{r}
library(caret)
```

## Prepare the Data

Load the raw data from local data folder into a dataset for training.

```{r load-data, cache=TRUE}
dsTraining <- read.csv("data//pml-training.csv", header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))

dim(dsTraining)
```

Eliminate columns where data is missing and those that don't have belt, forearm, arm, dumbbell data to find the predictors. Also, remove the highly correlated variables

```{r, dependson="load-data"}
isAnyMissing <- sapply(dsTraining, function (x) any(is.na(x) | x == ""))

isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(dsTraining))

predictors <- names(dsTraining)[isPredictor]

highlyCorrelated <- findCorrelation(cor(dsTraining[, predictors]), cutoff=0.9)
predictors <- predictors[-c(highlyCorrelated)]

predictors
```

Subset the dataset and split into 60% training and 40% testing set.

```{r}
dsTraining <- dsTraining[, c("classe", predictors)]

inTrain <- createDataPartition(y=dsTraining$classe, p=0.6, list=FALSE)
training <- dsTraining[inTrain,]
testing <- dsTraining[-inTrain,]

dim(training); dim(testing)
```

## Train the Model

Use a Random Forest method due to its popularity and accuracy to fit a model.  Use a seed to make the results reproducable.

```{r train-model, cache=TRUE, dependson="load-data"}
set.seed(12345)
trctrlRF <- trainControl(method="oob", number=3, repeats=1)
modFit <- train(classe ~ ., data=training, method="rf", 
                prox=TRUE, ntree=200, trControl=trctrlRF)
```

Check the accuracy against the training set.

```{r}
predictTrain <- predict(modFit, training)
confusionMatrix(predictTrain, training$classe)
```

Against the training dataset the model was able to predict the manner in which the exercise was done with 100% accuracy.  The 95% Confidence Interval (CI) is 99.97% to 100%.

Check the accuracy against the test set (cross-validation).

```{r}
predictTest <- predict(modFit, testing)
confusionMatrix(predictTest, testing$classe)
```

Against the testing dataset the model was able to predict the manner in which the exercise was done with 99.2% accuracy.  The 95% Confidence Interval (CI) is 98.97% to 99.38%.

## Perform the test

Load the raw data from local data folder into a dataset for testing.  Use the prediction model to find the answers for the manner in which the exercise was done.

```{r}
dsTest <- read.csv("data//pml-testing.csv", header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))

answers <- data.frame(predict(modFit, dsTest), dsTest$user_name)
answers
```

Generate the files for submission.

```{r}
pml_write_files = function(x){
  n = length(x)
  path <- "answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers[[1]])
list.files("answers")
```